\documentclass[main.tex]{subfiles}


\begin{document}


\section{Abstract}

RNA structures perform myriad functions, but analyzing structure ensembles remains challenging due to systematic biases in mutational profiling data.
Mutations less than four nucleotides apart are underrepresented, undermining the assumption of independent mutations and making single RNA structures appear as multiple clusters.
Here, we present SEISMIC-RNA, software that resolves RNA structure ensembles directly from raw sequencing reads while explicitly correcting for this bias and additional artifacts.
The fully automated workflow scales easily from single RNAs to whole transcriptomes; supports DMS-MaPseq, SHAPE-MaP, ETC-MaP, and other mutation-based assays; and outputs predicted structures plus 13 types of graphs for detailed analyses.
We benchmarked SEISMIC-RNA against existing software (DREEM, DRACO, and DanceMapper) using 4,320 simulated datasets spanning three RNA lengths and one to four true structures.
At 200,000 reads, SEISMIC-RNA correctly identified the number of clusters in 88\% of datasets, compared to 56–82\% for competing software, and ran 5–39 times faster.
It also pinpointed domains that form multiple structures in long transcripts using a new window-free algorithm that outperformed window-based approaches in our benchmarks.
Freely available and installable with a single Conda command, SEISMIC-RNA easily and accurately resolves RNA structure ensembles from complex mutational-profiling experiments.


\end{document}
