\documentclass[main.tex]{subfiles}


\begin{document}


\section{Acknowledgements}

We thank Yves J. Martin des Taillades, Albéric A. de Lajarte, and Scott L. Grote for contributions to the source code.
This research was supported by National Institute of Allergy and Infectious Diseases grant DP2 AI175475 (S.R.); National Science Foundation Graduate Research Fellowship Program grants 1745302 (M.F.A.), 2140743 (J.A.), and 2141064 (M.F.A.); and National Institute of General Medical Sciences grant T32 GM145407 (J.A.).


\section{Author Contributions}

S.R. and M.F.A. conceived the project.
M.F.A. developed the architecture and algorithms of SEISMIC-RNA.
J.A. implemented additional features of SEISMIC-RNA.
M.F.A. benchmarked SEISMIC-RNA and the other software.
M.F.A. wrote the manuscript.
All authors reviewed the manuscript and provided comments.


\section{Ethics Declarations}

The authors declare no competing interests.


\section{Data Availability}

Scripts for downloading data used in parameterization, simulating data for benchmarking, benchmarking each piece of software, and analyzing the results; simulated reference sequences and parameters; figures; and LaTeX source code for this manuscript are available from GitHub: \url{https://github.com/rouskinlab/seismic-rna-paper}.
Source code for SEISMIC-RNA is available from GitHub: \url{https://github.com/rouskinlab/seismic-rna}.
Documentation for SEISMIC-RNA, including instructions for installation and usage, is hosted on GitHub Pages: \url{https://rouskinlab.github.io/seismic-rna}.


\end{document}
